Subglacial and subaerial fluvial sediment transport capacity respond differently to water discharge variations

### 2024-03-18 Delaneyetal2024.tex
mauros edits moving forward

\usepackage[inline]{trackchanges} %for better track changes. finalnew option will compile document with changes incorporated.
\usepackage{soul}
\linenumbers
%%%%%%%
% As of 2018 we recommend use of the TrackChanges package to mark revisions.
% The trackchanges package adds five new LaTeX commands:
%
%  \note[editor]{The note}
%  \annote[editor]{Text to annotate}{The note}
%  \add[editor]{Text to add}
%  \remove[editor]{Text to remove}
%  \change[editor]{Text to remove}{Text to add}
%
% complete documentation is here: http://trackchanges.sourceforge.net/
%%%%%%%

@@

\newcommand{\aref}[1]{\textbf{Reference #1}}
\newcommand{\TODO}[1]{\textbf{TODO: \color{red}#1}}
\newcommand{\ian}[1]{{\textbf{\color{blue}Ian says:} \color{blue} #1} }
\newcommand{\mauro}[1]{{\textbf{\color{green}Mauro says:} \color{green} #1} }
\newcommand{\alpine}{\textit{ALPINE}\,}
\newcommand{\icesheet}{\textit{ICESHEET}\,}
\newcommand{\m}{$\,\mathrm{m}$\,}
\newcommand{\cm}{$\,\mathrm{cm}$\,}
\newcommand{\mma}{$\,\mathrm{mm  \, a^{-1}}$\,}
\newcommand{\mmma}{$\,\mathrm{m^3\, a^{-1}}$\,}
\newcommand{\mmms}{$\,\mathrm{m^3\, s^{-1}}$\,}
\newcommand{\unit}[1]{$\mathrm{#1}$}

@@

\affiliation{1}{Institut des dynamiques de la surface terrestre (IDYST), Universit\'{e} de Lausanne, B\^{a}timent G\'{e}opolis, 1015 Lausanne, Switzerland}
\affiliation{2}{Department of Geosciences, University of Fribourg, Ch. du Musée 1700, Fribourg, Switzerland}
\affiliation{3}{Laboratory of Hydraulics, Hydrology and Glaciology (VAW), ETH-Z\"urich, H\"onggerbergring 26, 8093 89 Z\"urich, Switzerland}
\affiliation{4}{Swiss Federal Institute for Forest, Snow and Landscape Research (WSL) Z\"uricherstrasse 111, 8903 1011 Birmensdorf, Switzerland}

@@

\begin{keypoints}
\item Short term water discharge variations in subglacial channels are mainly accommodated by water velocity, not channel size.
\item Greater variability in water velocity causes sediment transport capacity to vary more in subglacial channels than in subaerial channels.
\item Results suggest that significant sediment transport capacity can persist across a range of water discharge conditions, and timing of peak events occurs under different hydrological conditions.
\end{keypoints}

\begin{abstract}
  Sediment transport capacity in both subaerial and subglacial channels depends on the shear stress exerted across the channel bottom, which vary with the  water velocity and channel width.
  In subaerial channels, water discharge variations are accommodated largely by changes in flow depth and width.
  While in subglacial channels, water discharge changes primarily affect water velocity due to slow-evolving channel geometry.
  Hydrographs from an Alpine glacier and the Greenland Ice Sheet are applied to models of subglacial and subaerial hydraulics, revealing greater variability in the inferred sediment transport capacity in subglacial than in subaerial channels.
  Results also show that high subglacial sediment transport rates can occur across a wide range of water discharges.
  However, slowly evolving channel width accommodates some variability in subglacial channels, compared to subaerial channels.
  These findings are discussed in the context of understanding sediment discharge from glaciers with different hydrological forcings and assumptions of channel type.
\end{abstract}

\section*{Plain Language Summary} % 200 words
Sediment export from glaciers impacts downstream ecosystems and communities and it has been documented to change in response to climate warming and glacier retreat.
In these catchments, sediment export depends in part on the sediment transport capacity of water flowing below the glacier, or the amount of sediment that could be transported given the abundance of sediment.
This capacity largely depends on the width of the channel and the water's velocity. 
Unlike subaerial channels, like rivers, where the size of the river changes rapidly with respect to water discharge, however, subglacial channels are pressurized by the ice above.
This pressurized flow means that channel size evolves slowly compared to the variations in water discharge, therefore variations in water discharge are largely accommodated by variations in velocity.
Models of these processes show that sediment transport capacity is highly variable in subglacial channels, and a wide range of water discharges can result in comparable sediment transport capacity.
In subaerial channels, by contrast, sediment transport capacity is less variable and depends on a single water discharge quantity.
The channel's different response to water discharge may impact the interpretation of sediment transport records and responds to the type of hydrological forcing applied.

@@

Changes in glacier dynamics and hydrology  have prompted numerous  recent studies of  sediment transport processes in cold regions \cite<e.g.>{zhang2022}.
Increases in sediment transport  have been observed in Greenland \cite{bendixen2017}, the European Alps \cite{costa2017}, the Himalayas \cite{li2021}, and the Andes \cite{vergara2022}.
In some regions, increased water discharge and glacier melt have been interpreted to yield greater sediment transport capacity \cite{bendixen2017,costa2017,li2021}.
Observed changes to sediment transport in glacierized catchments require examining the processes controlling sediment discharge in these catchments and its variations with water discharge  \cite<e.g.>{riihimaki2005,swift2005}.

Over long periods, processes such as glacier abrasion and quarrying sculpt landscapes and create sediment to be transported fluvially from under glaciers \cite<c.f.>{hallet1979,iverson2012,ugelvig2018}.
On shorter time periods, pressurized water transports sediment from underneath glaciers \cite{walder1994,creyts2013,beaud2018,delaney2019}, should enough sediment be present subglacially (i.e. in a transport-limited regime).

In a transport-limited regime, sediment discharge tracks sediment transport capacity, defined as the amount of sediment the water can carry.
Sediment transport capacity depends on the shear stress between water and the sediment it flows  over \cite{shields1936,meyer1948,engelund1967} along with the width of the channel bottom $w$ over which to mobilize sediment.
The shear stress $\tau$ responds to the velocity of water $v$ flowing through the channel so that
\begin{linenomath*}
  \begin{equation}
    \label{eq:tau}
    \tau \propto v^2.
  \end{equation}
\end{linenomath*}
%
Following mass conservation, the mean velocity of the water flowing through a  channel is
\begin{linenomath*}
  \begin{equation}
    \label{eq:v}
    v = \frac{Q}{S},
  \end{equation}
\end{linenomath*}
where $Q$ is water discharge,  and $S$ is the channel's wetted area.

@@

The response of water velocity to changing water discharge in subglacial channels differs from subaerial ones.
The size of subglacial channels responds to the creep closure of the ice above and the opening of the channel by melting due to frictional heating of water flowing through the channel \cite{rothlisberger1972}.
As a result, the subglacial channel size only evolves relatively slowly over days to months, whereas water discharge can vary over hours \cite<e.g.>{iken1986,andrews2014,nanni2020}.
Therefore, on short time scales, subglacial water flow behaves more like pipe flow and changes in water discharge $Q$ are mainly accommodated with increased or decreased water velocity $v$ \cite{alley1997} because the size of the channel $S$ changes on a much longer timescale (Equation~\ref{eq:v} and Figure~\ref{fig:cartoon}).

  \begin{figure}[h]
  \centering
    \includegraphics[width=0.8\linewidth]{Fig1.pdf}
    \caption{Sketch for the different responses that subglacial and subaerial channels to increased water discharge.
      Water velocity magnitudes in the subglacial, $v_g$, and subaerial, $v_f$, channels are shown by arrow length.
      $S_g$ and $S_f$ represent the wetted area in subglacial and subaerial channels, respectively.
      The subglacial channel widths $w_g$ remain unchanged, while the subaerial channel widths $w_f$ evolve. Shear stress $\tau$ is responsible for the mobilization of sediment.
    }
    \label{fig:cartoon}
  \end{figure}

Consequently, sediment mobilization in subaerial and subglacial channels responds differently to changing water discharge.
These differences are implicitly included in a wide range of available models that  quantify sediment transport in both subglacial and subaerial channels  \cite<e.g.>{walder1994,alley1997,tucker1997,creyts2013,delaney2019,hewitt2019,wickert2019}.
The divergent response of changing water discharge on sediment mobilization capacity may well impact sediment dynamics at glacier margins where flow transitions from pressurized to open channel flow \cite<e.g.>{perolo2018}.
Additionally, the variable response of sediment transport capacity to water discharge in the two systems may affect the interpretation of sediment transport records in glacial systems \cite<e.g.>{muller1968,richards2003,swift2005,ganti2016}.
Yet, little work has examined the differing relationship between variations in water discharge and sediment discharge in subglacial and subaerial channels \cite{alley1997}.

This manuscript evaluates the relationship between sediment transport capacity and water discharge in subglacial channels and compares it to subaerial ones.
Models of subglacial and subaerial hydraulics are used to quantify subglacial and subaerial water velocity and channel morphology.
The models are applied to hydrological records from an Alpine glacier in Switzerland (Fieschergletscher) and  a land-terminating glacier in Greenland (Leverett Glacier).
The  lumped nature of the models isolates the disparity between water discharge and sediment transport capacity in subglacial systems, independent of the upstream drainage network and sediment access.
Outputs demonstrate differences in the relationships between  water discharge, channel geometry, and water velocity in the subglacial and subaerial channels.
The manuscript then uses the demonstrated differences between the two channel types to discuss the implications for interpreting the processes driving variability in records of sediment transport, and their relationship with hydrology from glacierized catchments.

@@

Water discharge data is leveraged from Alpine and ice sheet settings collected downstream of the glacier, assuming that no water storage occurred in the proglacial area.
The Alpine site (\alpine) is  Fieschergletscher in the Swiss Alps ($46^\circ\,29'\,07''$ N, $8^\circ\,08'\,3''$ E).
Water discharge was collected here at a $1$\unit{min} interval continuously over the period 2014--2021.
See \citeA{felix2022} and \citeA{felix2021} for more details.

The Leverett Glacier in Greenland (\icesheet) serves as the ice sheet setting.
Water discharge was collected roughly $2$\unit{km} downstream from the terminus between June and August/September 2009 to 2012 ($67^\circ\,03'\,5''$ N, $50^\circ\,12'\,59''$ W).
Data is used in \citeA{tedstone2017} and \citeA{hawkings2014} at a $5$\,\unit{min} time interval from May 28, 2012 to August 8, 2012.

\section{Methods}
\label{sect:meth}
The models described below (Sections ~\ref{sect:sub_mode}~and~\ref{sect:fluv}) represent relationships amongst water discharge, water velocity, and channel geometry in both subaerial and subglacial channels (Table \ref{table:vpm}).
Both models calculate   water velocity, shear stress, and width-integrated shear stress, upon which sediment transport depends \cite<Figure \ref{fig:cartoon}; >{shields1936}.
The evaluation of these variables omits the selection of a sediment transport relationship and a grain-size parameter \cite<e.g.>{shields1936,meyer1948} needed to calculate sediment transport capacity from shear stress.

@@

Here, it is assumed that the water is transported through a subglacial channel \cite<Figure~\ref{fig:cartoon}; >{rothlisberger1972}, and that the channel  size responds to melt due to frictional heating from water flow and to creep closure of the ice, using the Darcy-Weisbach formulation for water-flow through a pipe  \cite<e.g.>{rothlisberger1972,clarke2003}.
The formulation here does not consider the englacial storage of water.
The evolution of subglacial channel size $S_g$ is given as
\begin{linenomath*}
  \begin{equation}
    \label{eq:dS_dt}
    \frac{\partial S_g}{\partial t} = C_1 \frac{Q \Delta h}{l} - C_2 \left(h_{o}-\frac{\Delta h}{2}\right)^n\,S_g,
  \end{equation}
\end{linenomath*}
\noindent where $C_1= (1-\rho_wc_pc_t)\,\frac{\rho_wg}{\rho_iL}$ and $C_2=2A(\frac{\rho_wg}{n})^n$ are constants (Table \ref{table:vpm}), $l$ is the length of the subglacial channel, $Q$ is water discharge, $\Delta h$ is the hydraulic potential change from the glacier terminus to its top, $h_{o}= \frac{\rho_i}{\rho_w} h_{ice}$ is the ice overburden pressure expressed in meter water equivalent, and $n$ is Glen's n \cite{glen1955} (usually $n=3$).
The first term on the right side of the equation represents the opening of the channel through frictional heating, while the following term represents the creep closure of the channel from ice deformation.

Following the Darcy-Weisbach equation, the head drop $\Delta h$ is
\begin{linenomath*}
  \begin{equation}
    \label{eq:dh}
    \Delta h \,  = l \,\frac{1}{2g} \,f_r\,\frac{v_g^2}{D_h},
  \end{equation}
\end{linenomath*}
\noindent where $f_r$ is a friction factor, $D_h$ is the hydraulic diameter, $l$ is the length over which head change occurs, and $v_g=\frac{Q}{S_g}$ is the water flow speed.
%
The hydraulic diameter $D_h$ is converted to wetted area $S_g$ with
\begin{equation}
  \label{eq:Dh2S}
  S_g= \frac{D_h^2}{2}\,\frac{(\frac{\beta}{2}+\sin \frac{\beta}{2})^2}{\beta - \sin \beta},
\end{equation}
where $\beta$ is the central angle of the circular segment that comprises the channel (the so-called Hooke angle, \citeA{hooke1990}). Note that $\beta =\pi$ corresponds to a semi-circular channel and smaller values of $\beta$ result in shallow, wide channels.
%
The shear stress, $\tau_g$, between the water and the channel bed is determined through the Darcy-Weisbach formulation
\begin{equation}
  \label{eq:tau_g}
  \tau_g=\frac{1}{8}\,f_r\,\rho_w\,v_g^2,
\end{equation}
%
where $v_g = \frac{Q}{S_g}$ is the water velocity.
%
The width of the channel floor $w_g$ is represented as
\begin{equation}
  \label{eq:dh2wc}
  w_g = 2  \sin \frac{\beta}{2} \sqrt{\frac{2\, S_g}{\beta -\sin \beta}}.
\end{equation}
%
This value is used to establish the integrated shear stress across the channel, $w_g\tau_g$.

@@

Note that this subaerial channel model is purely algebraic, whereas the subglacial model comprises a differential equation for the evolution of $S_g$.
Thus the models have a history dependence on the discharge $Q_w$.

@@

The models above are applied to hydrological records from the Fieschergletscher (scenario \alpine) and the Leverett Glacier (scenario \icesheet).
The subglacial and subaerial models represent the subglacial flow of water exiting a glacier and transitioning to subaerial flow.
Outputs of the models represent generalizable sediment transport characteristics from these hydrographs, rather than actual hydraulic conditions.
To generalize these scenarios, \alpine{}  is exemplified by relatively thin ice thickness ($h_{ice}$: $225$\,\unit{m}), low water discharge ($\sim\,10$\,\unit{m}$^3$\,\unit{s}$^{-1}$) and high diurnal variability in water discharge (Figure~\ref{fig:model_outs}, a).
\icesheet{}  is exemplified by thick ice  ($h_{ice}$: $700$\,\unit{m}), high water discharge ($\sim\,300$\,\unit{m}$^3$\,\unit{s}$^{-1}$)  and low diurnal variability in water discharge (Figure~\ref{fig:model_outs}, e).
%
However, both scenarios share a discharge with a similar, relative seasonal amplitude.

For each glacier, the models are first applied to a reference test case for each glacier that assumes  a subglacial channel with $\beta=\frac{\pi}{6}$, i.e. a segment of a quarter-circle, (Table~\ref{table:vpm}).
Friction factors $f_r$ and $f_f$ are tuned so that reasonable water velocities occur for both \alpine and \icesheet{}\cite<$\sim\,1.6\,$\unit{m}$^3$\,\unit{s}$^{-1}$>{werder2010b,chandler2013}.
The subaerial channels of both scenarios have a bed slope of $\nabla z_c = 0.05$ (Table~\ref{table:vpm}).

A second experiment aims to further characterize the variability in sediment discharge capacity in subglacial compared to subaerial channels across a range channel shapes, friction factors and water discharge variability. 
To accomplish this, the water discharge records are aggregated across a select time period in the past ranging from $15$\,\unit{min} up to $15$\,\unit{days}, and the model is run with random  parameter combinations of channel geometry factors ($\beta$ and $k$) and friction factors ($f_r$ and $f_f$).
Runs with parameter combinations are accepted if their mean subglacial water velocity over the season lies between $0.75$\,\unit{m}$^3$\,\unit{s}$^{-1}$ and $2$\,\unit{m}$^3$\,\unit{s}$^{-1}$ or if subaerial water velocity lies between $0.3$\,\unit{m}$^3$\,\unit{s}$^{-1}$ and $1.2$\,\unit{m}$^3$\,\unit{s}$^{-1}$, in accordance with measurements \cite<e.g.>{werder2010b,magnusson2012,chandler2013}.
The routine runs until $100$ different parameter combinations are accepted from each water discharge aggregation period.
The variability of an output, a key difference between the subaerial and subglacial systems, is represented by  its  standard deviation over the course of the season.

@@

@@

In R-channel models, viable water pressure regimes can be achieved by changing the shape and friction factor of the channels, enabling assumptions of semi-circular channels in most applications \cite{werder2010}.
However, the role of channel width in the findings here suggests that more thorough parameterization of channel shape and channel width is needed in assessing subglacial sediment transport capacity (Section\ref{sect:sub_mode}, \ref{sect:scaling}).
Note as well that sediment transport relationships describe sediment flux per unit width, so that the integrated sediment transport capacity across channel occurs after application of the exponent, unlike the width-integrated shear stress evaluation here \cite<c.f.>{tucker1997}.

>>>>

\usepackage[inline]{trackchanges} %for better track changes. finalnew option will compile document with changes incorporated.
\usepackage{soul}
\linenumbers
%% make line numbers work with amsmath
% https://tex.stackexchange.com/questions/461186/how-to-use-lineno-with-amsmath-align
\usepackage{lineno}   %% <- no mathlines option
\usepackage{amsmath}  %% <- after lineno
\usepackage{etoolbox} %% <- for \cspreto, \csappto

%% Patch 'normal' math environments:
\newcommand*\linenomathpatch[1]{%
  \cspreto{#1}{\linenomath}%
  \cspreto{#1*}{\linenomath}%
  \csappto{end#1}{\endlinenomath}%
  \csappto{end#1*}{\endlinenomath}%
}

\linenomathpatch{equation}
\linenomathpatch{gather}
\linenomathpatch{multline}
\linenomathpatch{align}
\linenomathpatch{alignat}
\linenomathpatch{flalign}

%%%%%%%
% As of 2018 we recommend use of the TrackChanges package to mark revisions.
% The trackchanges package adds five new LaTeX commands:
%
%  \note[editor]{The note}
%  \annote[editor]{Text to annotate}{The note}
%  \add[editor]{Text to add}
%  \remove[editor]{Text to remove}
%  \change[editor]{Text to remove}{Text to add}
%
% complete documentation is here: http://trackchanges.sourceforge.net/
%%%%%%%

@@

\newcommand{\aref}[1]{\textbf{Reference #1}}
\newcommand{\TODO}[1]{\textbf{TODO: \color{red}#1}}
\newcommand{\ian}[1]{{\textbf{\color{blue}Ian says:} \color{blue} #1} }
\newcommand{\mauro}[1]{{\textbf{\color{green}Mauro says:} \color{green} #1} }
\newcommand{\alpine}{\textit{ALPINE}}
\newcommand{\icesheet}{\textit{ICESHEET}}
\newcommand{\m}{$\,\mathrm{m}$\,}
\newcommand{\cm}{$\,\mathrm{cm}$\,}
\newcommand{\mma}{$\,\mathrm{mm  \, a^{-1}}$\,}
\newcommand{\mmma}{$\,\mathrm{m^3\, a^{-1}}$\,}
\newcommand{\mmms}{$\,\mathrm{m^3\, s^{-1}}$\,}
\newcommand{\unit}[1]{$\mathrm{#1}$}

@@

\affiliation{1}{Institut des dynamiques de la surface terrestre (IDYST), Universit\'{e} de Lausanne, B\^{a}timent G\'{e}opolis, 1015 Lausanne, Switzerland}
\affiliation{2}{Department of Geosciences, University of Fribourg, Ch. du Musée 1700, Fribourg, Switzerland}
\affiliation{3}{Laboratory of Hydraulics, Hydrology and Glaciology (VAW), ETH-Z\"urich, H\"onggerbergring 26, 8093 Z\"urich, Switzerland}
\affiliation{4}{Swiss Federal Institute for Forest, Snow and Landscape Research (WSL) Z\"uricherstrasse 111, 8903 Birmensdorf, Switzerland}

@@

\begin{keypoints}
\item Short term water discharge variations in subglacial channels are mainly accommodated by water velocity, not channel size, as is the case in subaerial systems.
\item Greater variability in water velocity causes sediment transport capacity to vary more in subglacial channels than in subaerial channels.
\item Results suggest that significant subglacial sediment transport capacity can persist across a range of water discharge conditions, and the timing of peak events occurs under different hydrological conditions.
\end{keypoints}

\begin{abstract}
  Sediment transport capacity in both subaerial and subglacial channels depends on the shear stress exerted across the channel bottom, which varies with the  water velocity and channel width.
  In subaerial channels, water discharge variations are accommodated largely by changes in flow depth and width.
  While in subglacial channels, water discharge changes primarily affect water velocity due to evolving channel geometry over hours to weeks.
  Hydrographs from an Alpine glacier and the Greenland Ice Sheet are used to drive models of subglacial and subaerial hydraulics, revealing greater variability in the inferred sediment transport capacity in subglacial than in subaerial channels.
  Results also show that high subglacial sediment transport rates and water velocities can occur across a wide range of water discharges.
  These findings are discussed in the context of understanding sediment discharge from glaciers with different hydrological forcings and assumptions of channel type.
\end{abstract}

\section*{Plain Language Summary} % 200 words
Sediment export from glaciers impacts downstream ecosystems and communities and it has been documented to change in response to climate warming and glacier retreat.
In these catchments, sediment export depends in part on the sediment transport capacity, the amount of sediment that could be transported for the hydraulic conditions given the abundance of sediment.
This capacity largely depends on the width of the channel and the water's velocity. 
Unlike subaerial channels, like rivers, where the size of the river changes rapidly with water discharge, however, subglacial channels are pressurized by the ice above.
This pressurized flow means that channel size evolves slowly compared to the variations in water discharge, therefore variations in water discharge are largely accommodated by variations in velocity as opposed to channel size.
Models of these processes show that sediment transport capacity is highly variable in subglacial channels, and a wide range of water discharges can result in comparable sediment transport capacities.
In subaerial channels, by contrast, sediment transport capacity is less variable and depends solely on water discharge.
The channel's different responses to water discharge may impact the interpretation of sediment transport records and respond to the type of hydrological forcing applied.

@@

Changes in glacier dynamics and hydrology  have prompted numerous  recent studies of  sediment transport processes in cold regions \cite<e.g.>{zhang2022}.
Increases in sediment transport  have been observed in Greenland \cite{bendixen2017}, the European Alps \cite{costa2017}, the Himalayas \cite{li2021}, and the Andes \cite{vergara2022}.
In some regions, increased water discharge and glacier melt have been interpreted to yield greater sediment transport capacity \cite{bendixen2017,costa2017,li2021}.
To explain observed changes to sediment transport in glacierized catchments require examining the processes controlling sediment discharge in these catchments and its variations with water discharge  \cite<e.g.>{riihimaki2005,swift2005}.

Over long periods, processes such as glacier abrasion and quarrying sculpt landscapes and create sediment to be transported fluvially from under glaciers \cite<c.f.>{hallet1979,iverson2012,ugelvig2018}.
Pressurized subglacial water can transport this sediment from underneath glaciers \cite{walder1994,creyts2013,beaud2018,delaney2019}, should enough sediment be present subglacially (i.e. in a transport-limited regime).

In a transport-limited regime, sediment discharge tracks sediment transport capacity, defined as the amount of sediment the water can carry.
Sediment transport capacity depends on the shear stress between water and the sediment it flows  over \cite{shields1936,meyer1948,engelund1967} along with the width of the channel bottom $w$ over which sediment mobilizes.
The shear stress $\tau$ responds to the velocity of water $v$ flowing through the channel so that
\begin{linenomath*}
  \begin{equation}
    \label{eq:tau}
    \tau \propto v^2.
  \end{equation}
\end{linenomath*}
%
Following mass conservation, the mean velocity of the water flowing through a  channel is
\begin{linenomath*}
  \begin{equation}
    \label{eq:v}
    v = \frac{Q}{S},
  \end{equation}
\end{linenomath*}
where $Q$ is water discharge,  and $S$ is the channel's wetted area.

@@

The response of water velocity to changing water discharge in subglacial channels differs from subaerial ones.
The size of subglacial channels responds to the creep closure of the ice and the opening of the channel by melting due to frictional heating of water flowing through the channel \cite{rothlisberger1972}.
As a result, the subglacial channel size only evolves relatively slowly over days to months, whereas water discharge can vary over hours \cite<e.g.>{iken1986,andrews2014,nanni2020}.
Therefore, on short time scales, subglacial water flow behaves more like pipe flow and changes in water discharge $Q$ are mainly accommodated with increased or decreased water velocity $v$ \cite{alley1997} because the size of the channel $S$ changes on a much longer timescale (Equation~\ref{eq:v} and Figure~\ref{fig:cartoon}).

  \begin{figure}[h]
  \centering
    \includegraphics[width=0.8\linewidth]{Fig1.pdf}
    \caption{Sketch for the different short time-scale responses of subglacial and subaerial channels to increased water discharge.
      Water velocity magnitudes in the subglacial, $v_g$, and subaerial, $v_f$, channels are shown by arrow length.
      $S_g$ and $S_f$ represent the wetted area in subglacial and subaerial channels, respectively.
      The subglacial channel width $w_g$ remain unchanged, while the subaerial channel width $w_f$ evolves. Shear stress $\tau$ is responsible for the mobilization of sediment.
    }
    \label{fig:cartoon}
  \end{figure}

Consequently, sediment mobilization in subaerial and subglacial channels responds differently to changing water discharge.
These differences are implicitly included in a wide range of available models that  quantify sediment transport in both subglacial and subaerial channels  \cite<e.g.>{walder1994,alley1997,tucker1997,creyts2013,delaney2019,hewitt2019,wickert2019}.
The divergent response of changing water discharge on sediment mobilization capacity may well impact sediment dynamics near glacier margins where flow transitions from pressurized to open channel flow \cite<e.g.>{perolo2018}.
Additionally, the variable response of sediment transport capacity to water discharge in the two systems may affect the interpretation of sediment transport records in glacial systems \cite<e.g.>{muller1968,richards2003,swift2005,ganti2016}.
Yet, little work has examined the differing relationship between variations in water discharge and sediment discharge in subglacial and subaerial channels \cite{alley1997}.

This manuscript evaluates the relationship between sediment transport capacity and water discharge in subglacial channels and compares it to subaerial ones.
Models of subglacial and subaerial hydraulics are used to quantify the respective water velocity and channel morphology.
The models are applied to hydrological records from an Alpine glacier in Switzerland (Fieschergletscher) and  a land-terminating glacier in Greenland (Leverett Glacier).
The  lumped nature of the models isolates the disparity between water discharge and sediment transport capacity in subglacial systems, independent of the upstream drainage network and sediment access.
Outputs demonstrate differences in the relationships between  water discharge, channel geometry, and water velocity in the subglacial and subaerial channels.
The manuscript then uses the demonstrated differences between the two channel types to discuss the implications for interpreting the processes driving variability in records of sediment transport, and their relationship with hydrology from glacierized catchments.

@@

Water discharge data is used from Alpine and ice sheet settings collected downstream of the glacier, assuming that no water storage occurred in the proglacial area.
The Alpine site (\alpine{}) is  Fieschergletscher in the Swiss Alps ($46^\circ\,29'\,07''$ N, $8^\circ\,08'\,3''$ E).
Water discharge was collected here at a $1$\unit{min} interval continuously over the period 2014--2021;
see \citeA{felix2022} and \citeA{felix2021} for more details.

The Leverett Glacier in Greenland (\icesheet{}) serves as the ice sheet setting.
Water discharge was collected roughly $2$\unit{km} downstream from the terminus between June and August/September 2009 to 2012 ($67^\circ\,03'\,5''$ N, $50^\circ\,12'\,59''$ W).
Data is used in \citeA{tedstone2017} and \citeA{hawkings2014} at a $5$\,\unit{min} time interval from May 28, 2012 to August 8, 2012.

\section{Methods}
\label{sect:meth}
The models described below (Sections ~\ref{sect:sub_mode}~and~\ref{sect:fluv}) represent relationships amongst water discharge, water velocity, and channel geometry in both subaerial and subglacial channels (Table \ref{table:vpm}).
Both models use the measured discharge to calculate water velocity, shear stress, and width-integrated shear stress, upon which sediment transport depends \cite<Figure \ref{fig:cartoon}; >{shields1936}.
The evaluation of these variables omits the selection of a sediment transport relationship and a grain-size parameter \cite<e.g.>{shields1936,meyer1948} needed to calculate sediment transport capacity from shear stress.

@@

Here, it is assumed that the water is transported through a subglacial channel \cite<Figure~\ref{fig:cartoon}; >{rothlisberger1972}, and that the channel  size responds to melt due to frictional heating from water flow and to creep closure of the ice, using the Darcy-Weisbach formulation for water-flow through a pipe  \cite<e.g.>{rothlisberger1972,clarke2003}.
The formulation here does not consider the englacial storage of water.
The evolution of subglacial channel size $S_g$ is given as
\begin{linenomath*}
  \begin{equation}
    \label{eq:dS_dt}
    \frac{\partial S_g}{\partial t} = C_1 \frac{Q \Delta h}{l} - C_2 \left(h_{o}-\frac{\Delta h}{2}\right)^n\,S_g,
  \end{equation}
\end{linenomath*}
\noindent where $C_1= (1-\rho_wc_pc_t)\,\frac{\rho_wg}{\rho_iL}$ and $C_2=2A(\frac{\rho_wg}{n})^n$ are constants (Table \ref{table:vpm}), $g$ is the acceleration due to gravity, $Q$ is water discharge, $\Delta h$ is the hydraulic potential change over $l$, $h_{o}= \frac{\rho_i}{\rho_w} h_{ice}$ is the mean ice overburden pressure expressed in meter water equivalent, and $n$ is Glen's n \cite{glen1955} (usually $n=3$).
The first term on the right side of the equation represents the opening of the channel through frictional heating, while the following term represents the creep closure of the channel from ice deformation.

Following the Darcy-Weisbach equation, the head drop $\Delta h$ is
\begin{linenomath*}
  \begin{equation}
    \label{eq:dh}
    \Delta h \,  = l \,\frac{1}{2g} \,f_r\,\frac{v_g^2}{D_h},
  \end{equation}
\end{linenomath*}
\noindent where $f_r$ is a friction factor, $D_h$ is the hydraulic diameter, $l$ is the length over which head change occurs, and $v_g=\frac{Q}{S_g}$ is the water flow speed.
%
The hydraulic diameter $D_h$ is converted to wetted area $S_g$ with
\begin{equation}
  \label{eq:Dh2S}
  S_g= \frac{D_h^2}{2}\,\frac{(\frac{\beta}{2}+\sin \frac{\beta}{2})^2}{\beta - \sin \beta},
\end{equation}
where $\beta$ is the central angle of the circular segment that comprises the channel (the so-called Hooke angle, \citeA{hooke1990}). Note that $\beta =\pi$ corresponds to a semi-circular channel and smaller values of $\beta$ result in shallow, wide channels.
This completes the subglacial hydraulic model which is described by the state variables $S_g$ and $\Delta h$.

The shear stress, $\tau_g$, between the water and the channel bed is determined through the Darcy-Weisbach formulation
\begin{equation}
  \label{eq:tau_g}
  \tau_g=\frac{1}{8}\,f_r\,\rho_w\,v_g^2,
\end{equation}
%
where $v_g = \frac{Q}{S_g}$ is the water velocity and $\rho_w$ the density of water.
%
The width of the channel floor $w_g$ is represented as
\begin{equation}
  \label{eq:dh2wc}
  w_g = 2  \sin \frac{\beta}{2} \sqrt{\frac{2\, S_g}{\beta -\sin \beta}}.
\end{equation}
%
This value is used to establish the integrated shear stress across the channel $w_g\tau_g$.

@@

Note that this subaerial channel model is purely algebraic, whereas the subglacial model comprises a differential equation for the evolution of $S_g$.
Thus the subglacial model has a history dependence on the discharge $Q_w$.

@@

The models above are applied to proglacial discharge records from the Fieschergletscher (scenario \alpine{}) and the Leverett Glacier (scenario \icesheet{}).
The subglacial and subaerial models represent the subglacial flow of water exiting a glacier and transitioning to subaerial flow.
Outputs of the models represent generalizable sediment transport characteristics from these hydrographs, rather than actual hydraulic conditions.
To generalize these scenarios, \alpine{}  is exemplified by relatively thin ice thickness ($h_{ice}$: $225$\,\unit{m}), low water discharge ($\sim\,10$\,\unit{m}$^3$\,\unit{s}$^{-1}$) and high diurnal variability in water discharge (Figure~\ref{fig:model_outs}, a).
\icesheet{}  is exemplified by thick ice  ($h_{ice}$: $700$\,\unit{m}), high water discharge ($\sim\,300$\,\unit{m}$^3$\,\unit{s}$^{-1}$)  and low diurnal variability in water discharge (Figure~\ref{fig:model_outs}, e).
%
However, both scenarios share a discharge with a similar, relative seasonal amplitude.

For each glacier, the models are first applied to a reference test case for each glacier that assumes  a subglacial channel with $\beta=\frac{\pi}{6}$, i.e. a segment of a quarter-circle, (Table~\ref{table:vpm}).
Both friction factors $f_r$ and $f_f$ are tuned so that reasonable water velocities occur for both \alpine{} and \icesheet{} \cite<$\sim\,1.6\,$\unit{m}$^3$\,\unit{s}$^{-1}$>{werder2010b,chandler2013}.
The initial condition of cross-sectional area $S_g$ is evaluated by applying a model spinup with the maximum water discharge for the first $3$ days of the study until the change in channel area is negligible. 
The subaerial channels of both scenarios have a bed slope of $\nabla z_c = 0.05$ (Table~\ref{table:vpm}).

@@

A second experiment aims to further characterize the variability in sediment discharge capacity in subglacial compared to subaerial channels across a range of channel shapes, friction factors, and water discharge variability. 
The variability of one model output (i.e water velocity), a key difference between the subaerial and subglacial systems, is represented by its standard deviation throughout the season.

The effect of channel shape and roughness is established by running the model with random parameter values of channel geometry factors ($\beta$ and $k$) and friction factors ($f_r$ and $f_f$).
Runs with parameter combinations are accepted if their mean subglacial water velocity over the season lies between $0.75$\,\unit{m}$^3$\,\unit{s}$^{-1}$ and $2$\,\unit{m}$^3$\,\unit{s}$^{-1}$ or if subaerial water velocity lies between $0.3$\,\unit{m}$^3$\,\unit{s}$^{-1}$ and $1.2$\,\unit{m}$^3$\,\unit{s}$^{-1}$, following measurements \cite<e.g.>{werder2010b,magnusson2012,chandler2013}.
Additionally, to be accepted, model runs can only experience a flotation fraction of above $1.2$ for $2.5$\,\% of the run.

To establish the role of water discharge variability in driving variations in sediment transport capacity, the water discharge records are aggregated across a select period ranging from $15$\,\unit{min} up to $15$\,\unit{days}.
Aggregation periods longer than this might be conflated with seasonal variations in water discharge and are thus omitted.
In this case, the water discharge input at a given time is the average over the aggregation period before that time.  
The routine runs until $100$ different parameter combinations are accepted for each water discharge aggregation period.

@@

In R-channel models, viable water pressure regimes can be achieved by changing the shape and friction factor of the channels, enabling assumptions of semi-circular channels in most applications \cite{werder2010b}.
However, the role of channel width in the findings here suggests that more thorough parameterization of channel shape and channel width is needed in assessing subglacial sediment transport capacity (Section\ref{sect:sub_mode}, \ref{sect:scaling}).
\ian{Similar feedback occurs in subaerial channels \cite{phillips2016}, albeit at  timescales longer than days in subglacial channels. }
